---
title: "Postgress to R Examples"
author: "Daniel Hocking"
date: "January 20, 2015"
output: html_document
---

This document describes and gives examples of pulling data from the conte PostgreSQL database on our server (Felek). For relatively simple query's `dplyr` can be used. For more complex queries and spatial joins the R package `RPostgreSQL` is best.

## Get list of unique catchments with daymet data in our database using `RPostgreSQL`

```{r, eval=FALSE}
library(RPostgreSQL)
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, dbname="conte_dev", host="127.0.0.1", user="...", password="...")
qry <- "SELECT DISTINCT featureid FROM daymet;"
result <- dbSendQuery(con, qry)
catchments <- fetch(result, n=-1)
```

To access the database from outside a Felek RStudio session (i.e. from your local computer) you can use `con <- dbConnect(drv, dbname="conte_dev", host="felek.cns.umass.edu", user="...", password="...")`. Using `n = -1` in the `fetch` function means return all rows.

## Create a query that lists all locations and creates 'site' column using `dplyr`

```{r, eval=FALSE}
  qry_locations <- tbl(db, 'locations') %>%
    rename(location_id=id, location_name=name, featureid=catchment_id) %>%
    select(location_id, location_name, latitude, longitude, featureid, agency_id) %>%
    left_join(tbl(db, 'agencies') %>% rename(agency_name=name, agency_id=id) %>% select(agency_name, agency_id)) %>%
    select(location_id, location_name, featureid, latitude, longitude, agency_name) %>%
    mutate(site=concat(agency_name, '_', location_name))
```

## Get latitude and longitude centroids for each catchment by featureid using `RPostgreSQL`

This uses the `ST_Y`, `ST_X`, and `ST_Centroid` functions within Postgres.

```{r, eval=FALSE}
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, dbname="conte_dev", host="127.0.0.1", user="...", password="...")
rs <- dbSendQuery(con, "SELECT featureid, 
                  ST_Y(ST_Centroid(geom)) as latitude, 
                  ST_X(ST_Centroid(geom)) as longitude 
                  FROM catchments;")
featureid_lat_lon <- fetch(rs, n=-1) 
```
